---
title: "Figures 1C-D: Frequency distribution of variants in MSAs"
output: html_notebook
---

This notebook creates plots from pre-computed missense variant frequencies
over MSA columns in the Pfam-gnomAD dataset.

The first plot shows the frequency distributions of the number of missense
variants in Pfam MSA columns in Pfams with different numbers of human sequences.

The second plot shows the relationship between the number of missense and
synonymous variants in an MSA column and the Shenkin diversity in SH2 domains.


```{r Setup and data processing}
suppressPackageStartupMessages(library(tidyverse))


# Load and process dataset
pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv <- read_csv("~/Library/CloudStorage/OneDrive-UniversityofDundee/Projects/VarAlign-2022/data/pfam-gnomAD-clinvar-pdb-colstats_c7c3e19.csv.gz",
comment = "#", show_col_types = FALSE)
avs <- pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv  # avs: alignments variants structure

# Filter unneccesary columns
whitelist <- c(
  "family",
  "column",
  "splice_region_variant&synonymous_variant",
  "synonymous_variant",
  "occupancy",
  "shenkin",
  "_missense_all",
  "mes_or",
  "mes_p",
  "shenkin_nrank",
  "mes_or_log"
)
blacklist <- !names(avs) %in% whitelist
avs[, blacklist] <- NULL

# Fix variable name
names(avs)[names(avs) == '_missense_all'] <- 'missense_all'

# Calculate total synonymous count
avs[, 'synonymous_all'] <- avs[, 'splice_region_variant&synonymous_variant'] +
  avs[, 'synonymous_variant']

# Filters
avs <- avs |> filter(occupancy > 0)  # must have human sequences

# Clear raw dataset from memory
rm(pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv)
```


```{r Figure 1C: Column missense frequency histograms by N human sequences}
(avs
 |> filter(occupancy %in% c(1, 5, 9))  # N human sequences to be plotted, filters columns
 # |> sample_n(10000)  # for prototyping
 |> group_by(occupancy, missense_all)
 |> summarise(n_cols = n())
 -> avs_missense_freqs_by_nhuman
 )

(avs_missense_freqs_by_nhuman
  # data
  |> filter(occupancy %in% c(1, 5, 9))
  
  # plot
  |> ggplot()
  + aes(x = missense_all, y = n_cols)
  + geom_bar(stat = 'identity')
  + facet_wrap(~ occupancy, scales = 'free_y')
  
  # annotation
  + xlab('# Missense')
  # + ylab('# Pfam Columns')
  + scale_y_continuous(name = '# Pfam Columns', labels = scales::comma)
  
  # plot theme
  + theme_minimal()
  + theme(aspect.ratio = 1)
  + theme(
    axis.line = element_line(color='black'),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(size = 0.1),
    panel.grid.major.x = element_blank(),
    
    # hide facet labels
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    )
  )

ggsave('plot1c.svg')
```



```{r Figure 1D: Column variant totals vs. Shenkin in SH2 domains}

# Graph to help pick occupancy threshold for the figure
# This is required because Shenkin is strongly dependent on occupancy in the
# range from 1 - ~100, which applies to extremely gappy regions found in Pfams.
# n.b. The regression model handles this specifically but here I think its best
# to filter these artefacts to focus on the informative region of the data.
(avs
 |> filter(family == 'PF00017')
 |> ggplot()
 + aes(x = occupancy, y = shenkin)
 + geom_point()
 + geom_vline(xintercept = 90, lty = 2)
 + theme_minimal()
 + theme(aspect.ratio = 1)
 + theme(axis.line = element_line(color='black'))
)

(avs
 |> filter(family == 'PF00017')
 |> filter(occupancy >= 90)
 |> select(column, occupancy, shenkin, missense_all, synonymous_all)
 |> pivot_longer(missense_all:synonymous_all, names_to = 'variant', values_to = 'n')
  
  # Plot
 |> ggplot()
 + aes(x = shenkin, y = n)
 + facet_grid(~ variant)
 + geom_smooth(formula = y ~ x, method = 'lm', colour = 'black')
 + geom_point()
  
  # Annotation
  + scale_x_continuous(name = 'Shenkin', breaks = c(0, 25, 50, 75, 100),
                       labels = c(0, 25, 50, 75, 100), limits = c(0, 100))
  + ylab('# Missense')
 
  # Theme
  + theme_minimal()
  + theme(aspect.ratio = 1)
  + theme(
    axis.line = element_line(color='black'),  # heavy line for axes
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    
    # hide facet labels
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    )
 )
ggsave('plot1d.svg')
```

```{r Help select threhold for MES colouring in Jalview to make Figure 1F}
(avs
 |> filter(family == 'PF00017')
 |> group_by(occupancy >= 90, mes_or < 1, mes_p < 0.1)
 |> summarise(max(mes_or_log))
 )
```

