---
title: "Figure 3A: The Conservation Plane"
output:
  github_document:
    toc: true
    toc_depth: 2
---

# Introduction
This R markdown document visualizes the relationship between evolutionary
conservation and population constraint in Pfam domains. The dataset used for
this analysis is obtained from our pre-calculated MSA column statistics for
Pfam, included in this repository. The resulting plot corresponds to Figure 3A
in the manuscript.

# Notebook resources
## Libraries

```{r Libraries and globals}
suppressPackageStartupMessages(library(tidyverse))


COLOUR_SCHEME <- c(
  'cmd' = '#CC79A7',
  'ume' = '#56B4E9',
  'umn' = '#999999',
  'cmn' = '#999999',
  'umd' = '#E69F00',
  'cme' = '#009E73'
)
```

## Dataset

```{r Load dataset}
# Read and process dataset
pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv <-
  read_csv(
    "data/pfam-gnomAD-clinvar-pdb-colstats_c7c3e19.csv.gz",
    comment = "#",
    show_col_types = FALSE
  )

# Processing
avs <-
  pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv  # avs: alignments variants structure

# Filter unnecessary variables
whitelist <- c(
  "family",
  "column",
  "splice_region_variant&synonymous_variant",
  "synonymous_variant",
  "occupancy",
  "shenkin",
  "_missense_all",
  "mes_or",
  "mes_p",
  "occupancy_minus_threshold",
  "occupancy_gteq_threshold",
  "shenkin_nrank",
  "mes_or_nrank",
  "mes_or_log",
  "PDB_jury_column_rsa_unb"
)
blacklist <- !names(avs) %in% whitelist
avs[, blacklist] <- NULL

# Better variable name format
names(avs)[names(avs) == '_missense_all'] <- 'missense_all'

# Calculate synonymous all
avs[, 'synonymous_all'] <-
  avs[, 'splice_region_variant&synonymous_variant'] +
  avs[, 'synonymous_variant']

# Clean-up
rm(pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv)
```

```{r Categorise columns}
# Apply thresholds to categorise MES and Shenkin and the combined class
avs <- (
  avs
  |> mutate(mes_class = factor(
    ifelse(mes_p < 0.1, ifelse(mes_or < 1, 'md', 'me'), 'mn'), levels = c('md', 'mn', 'me')
  ))
  |> mutate(cons_class = factor(
    ifelse(shenkin_nrank > 0.5, 'u', 'c'), levels = c('u', 'c')
  ))
  |> mutate(xclass = factor(
    interaction(cons_class, mes_class, sep = ''),
    levels = c('cmd', 'cmn', 'cme', 'umd', 'umn', 'ume')
  ))
)
```

```{r Standard filters}
# Minimal filters for notebook
avs <- (
  avs
  # Column filters
  |> filter(occupancy > 0)  # must have human sequences
)
```

# Plot sample
```{r Sample the full dataset before plotting}
# We will plot a sample of the data so setting a seed to ensure reproducibility
set.seed(777)
(mes_vs_cons_data <- 
  avs
  |> filter(occupancy_gteq_threshold)
  |> filter(occupancy > 20)
  |> sample_n(10000)  # good balance for processing, completeness and legibility
  |> mutate(
    mes_class = factor(
      ifelse(mes_p < 0.1, ifelse(mes_or < 1, 'md', 'me'), 'mn'),
      levels = c('md', 'mn', 'me')
    ),
    cons_class = factor(
      ifelse(shenkin_nrank > 0.5, 'u', 'c'),
      levels = c('u', 'c')
    ),
    xclass = factor(
      interaction(cons_class, mes_class, sep = ''),
      levels = c('cmd', 'cmn', 'cme', 'umd', 'umn', 'ume')
    )
  )
)
```

```{r Summarise the sample}
mes_vs_cons_data |>
  summarise(n(), n_distinct(family))

mes_vs_cons_data |>
  group_by(PDB_jury_column_rsa_unb) |>
  summarise(n(), n_distinct(family))
```

# Figure 3A: The Conservation Plane (manuscript figure)
```{r Plot of the conservation plane}
mes_vs_cons_data |>
  arrange(mes_p) |>  # points with lower p will be plotted on top
  ggplot() +
  aes(x = shenkin_nrank, y = mes_or) +
  geom_point(aes(colour = xclass, alpha = ifelse(mes_p < 0.1, 1-mes_p, 0.01))) +
  scale_alpha_continuous(range = c(0.1, 0.7)) +
  scale_color_manual(values = COLOUR_SCHEME) +
  geom_hline(yintercept = 1, lty = 2) +
  geom_vline(xintercept = 0.5, lty = 2) +
  
  # Annotation
  # https://stackoverflow.com/questions/28709331/logarithmic-grid-for-plot-with-ggplot2
  scale_y_log10(breaks = 10^(-10:10),
                minor_breaks = rep(1:9, 21)*(10^rep(-10:10, each=9)),
                labels = 10^(-10:10)) +
  annotation_logticks(sides = 'l') +
  xlab('Normalised Shenkin') +
  ylab('MES') +
  annotate("text", x = 0.4, y = 1.20, label = "CME") +
  annotate("text", x = 0.4, y = 0.83, label = "CMD") +
  annotate("text", x = 0.6, y = 1.20, label = "UME") +
  annotate("text", x = 0.6, y = 0.83, label = "UMD") +

  # Plot theme
  theme_minimal() +
  theme(legend.position = "none", aspect.ratio = 1) +
  theme(
    axis.line = element_line(color='black'),  # heavy line for axes
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    # Hide facet labels
    strip.background = element_blank(),
    strip.text.x = element_blank(),
    ) ->
  mes_vs_cons_plot

mes_vs_cons_plot
```

```{r Save figure, include=FALSE}
# Save the plot
ggsave('plot3a.svg', mes_vs_cons_plot)
ggsave('plot3a.png', mes_vs_cons_plot, dpi = 300)
```

