---
title: "Figure 4B: Enrichment of protein-protein interactions"
output:
  github_document:
    toc: true
    toc_depth: 2
---

# Notebook resources
## Libraries

```{r Libraries}
suppressPackageStartupMessages(library(tidyverse))
library(patchwork)
```

## Dataset

```{r Load dataset}
# Read and process dataset
pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv <- read_csv("data/pfam-gnomAD-clinvar-pdb-colstats_c7c3e19.csv.gz",
comment = "#")

avs <- pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv  # avs: alignments variants structure

# Filter unnecessary variables
whitelist <- c(
  "family",
  "column",
  "occupancy",
  "shenkin",
  "_missense_all",
  "mes_or",
  "mes_p",
  "occupancy_minus_threshold",
  "occupancy_gteq_threshold",
  "shenkin_nrank",
  "mes_or_nrank",
  "mes_or_log",
  "PDB_sequences_with_contacts",
  "PDB_protein_protein_interactions",
  "PDB_total_protein_interactions",
  "PDB_non_self_ppis",
  "PDB_total_non_self_ppis",
  "PDB_surface_ppis",
  "PDB_total_surface_ppis",
  "PDB_surface_homo_ppis",
  "PDB_total_surface_homo_ppis",
  "PDB_homo_ppis",
  "PDB_total_homo_ppis",
  "PDB_intra_domain_x_interactions",
  "PDB_total_intra_domain_x_interactions",
  "PDB_intra_domain_domain_interactions",
  "PDB_total_intra_domain_domain_interactions",
  "PDB_jury_column_rsa_unb",
  "PDB_jury_column_rsa_bound"
)
blacklist <- !names(avs) %in% whitelist
avs[, blacklist] <- NULL

# Better variable name format
names(avs)[names(avs) == '_missense_all'] <- 'missense_all'

avs <- (
  avs
  
  # Categorise MES and Shenkin
  |> mutate(mes_class = factor(
    ifelse(mes_p < 0.1, ifelse(mes_or < 1, 'md', 'me'), 'mn'), levels = c('md', 'mn', 'me')
  ))
  |> mutate(cons_class = factor(
    ifelse(shenkin_nrank > 0.5, 'u', 'c'), levels = c('u', 'c')
  ))
  |> mutate(xclass = factor(
    interaction(cons_class, mes_class, sep = ''),
    levels = c('cmd', 'cmn', 'cme', 'umd', 'umn', 'ume')
  ))
  
  # Other
  |> mutate(PDB_jury_column_rsa_unb = as.factor(PDB_jury_column_rsa_unb))  #include all RSA/MES intersections
  |> replace_na(list(PDB_protein_protein_interactions = 0,
                     PDB_surface_ppis = 0,
                     PDB_sequences_with_contacts = 0))
)


# Clean-up
rm(pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv)


COLOUR_SCHEME <- c(
  'cmd' = '#CC79A7',
  'ume' = '#56B4E9',
  'umn' = '#999999',
  'cmn' = '#999999',
  'umd' = '#E69F00',
  'cme' = '#009E73'
)
```


# Calculate and plot odds ratios

1. Filter Pfams without any protein-protein interactions
2. Filter low-powered Pfams
3. Calculate summary statistics for each X-Class
4. Calculate odds ratios for each X-class
5. Repeat steps 3-4 for individual RSA classes
6. Compose plot

First we apply the filters:

```{r Filter dataset for these tests}
# Summarise counts
avs_test_set <- (
  avs
  
  # Column filters
  |> filter(occupancy > 0)  # must have human sequences
  |> filter(PDB_sequences_with_contacts > 0)  # drop columns with no PDB data
  
  # Pfam filters
  |> group_by(family) |> filter(any(PDB_protein_protein_interactions > 0)) |> ungroup()
  
  # Heuristic to select well-powered Pfams
  |> group_by(family) |> filter(any(mes_or < 1 & mes_p < 0.1)) |> ungroup()  # at least one depleted site
  |> group_by(family) |> filter(any(mes_or > 1 & mes_p < 0.1)) |> ungroup()  # at least one enriched site
)
```

Then summarise the data over the X-Classes:

```{r Summary statistics for different X-Classes}
summarize_data <- function(data, group_vars) {
  data |>
    group_by(across(all_of(group_vars))) |>
    summarise(
      n_pfams = n_distinct(family),
      n_columns = n(),
      n_residues = sum(occupancy),
      n_missense = sum(missense_all),
      across(c(PDB_protein_protein_interactions, PDB_sequences_with_contacts), sum),
      cons_class = first(cons_class),
      mes_class = first(mes_class)
    )
}

add_ratios <-
  function(data) {
    data %>%
      mutate(
        missense_per_res = n_missense / n_residues,
        ppi_per_pdb = PDB_protein_protein_interactions / PDB_sequences_with_contacts,
        PDB_per_column = PDB_sequences_with_contacts / n_columns
      )
  }

add_test_backgrounds <- function(data, group_vars) {
  data |>
    ungroup() |>
    group_by(across(all_of(group_vars))) |>
    mutate(across(
      c(n_missense, n_residues, PDB_protein_protein_interactions, PDB_sequences_with_contacts),
      sum,
      .names = "test_background_{.col}"
    ))
}


# Calculate group statistics for different X-Classes
avs_xclass_summary <- avs_test_set |>
  summarize_data(group_vars = "xclass") |>
  add_ratios() |>
  add_test_backgrounds(group_vars = NULL)
avs_xclass_summary
```

```{r Totals}
# Totals
avs_xclass_summary |> ungroup() |> select(where(is.numeric)) |> summarise_all(sum)
```

```{r Are protein-protein interactions associated with conservation plane class?}
# Chi-square test
chi <- chisq.test(avs_xclass_summary[, c("PDB_protein_protein_interactions", "PDB_sequences_with_contacts")])
chi
```
## Odds ratios for protein-protein interaction enrichment in X-Classes
Now apply Fishers exact test for each X-Class (CMD, UMD, etc...).

```{r Test enrichment of protein-protein interactions in different X-Classes}
perform_fisher_test <- function(df, group_vars) {
  df %>%
    group_by(!!!syms(group_vars)) %>%
    summarise(
      test = broom::tidy(
        fisher.test(matrix(
          c(
            PDB_protein_protein_interactions,
            PDB_sequences_with_contacts,
            test_background_PDB_protein_protein_interactions - PDB_protein_protein_interactions,
            test_background_PDB_sequences_with_contacts - PDB_sequences_with_contacts
          ),
          nrow = 2
        ), alternative = 'two.sided')
      ),
      cons_class = first(cons_class),
      mes_class = first(mes_class)
    ) %>%
    unnest(test)
}

avs_xclass_fisher <- perform_fisher_test(avs_xclass_summary, "xclass")
avs_xclass_fisher
```


Now do the same while stratifying for RSA class as well.

```{r Summarise data}
avs_xclass_summary_byRSA <- avs_test_set |>
  filter(PDB_jury_column_rsa_unb %in% c('Core', 'Part. Exposed', 'Surface')) |>
  summarize_data(group_vars = c("PDB_jury_column_rsa_unb", "xclass")) |>
  add_ratios() |>
  add_test_backgrounds(group_vars = "PDB_jury_column_rsa_unb")
avs_xclass_summary_byRSA
```

```{r Totals by RSA}
# Totals
avs_xclass_summary_byRSA |> ungroup() |> select(where(is.numeric)) |> summarise_all(sum)
```

## Odds ratios for protein-protein interaction enrichment in X-Class * RSA

```{r Test enrichment of protein-ligand interactions in X-Classes by RSA}
avs_xclass_fisher_byRSA <- perform_fisher_test(avs_xclass_summary_byRSA, c("PDB_jury_column_rsa_unb", "xclass"))
avs_xclass_fisher_byRSA
```
  
  
# Figure 4B: Odds ratio plot

```{r Combined plot}
plot_odds_ratio <- function(df, facet_var = NULL) {
  dodge_width <- 0.6
  
  plot <- df |>
    ggplot() +
    aes(x = xclass, y = estimate, colour = xclass) +
    geom_hline(yintercept = 1, lty = 2) +
    geom_errorbar(
      aes(ymin = conf.low, ymax = conf.high),
      width = 0.5,
      position = position_dodge(dodge_width)
    ) +
    geom_point(position = position_dodge(dodge_width)) +
    scale_y_log10() +
    ylab('Odds Ratio') +
    xlab('') +
    theme_minimal() +
    theme(legend.position = "none", aspect.ratio = 1) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) +
    scale_color_manual(values = COLOUR_SCHEME)
  
  if (!is.null(facet_var)) {
    plot <- plot + facet_grid(paste0(" ~ ", facet_var))
  }
  
  return(plot)
}


feature_oddsratio_plot <- plot_odds_ratio(avs_xclass_fisher)
feature_oddsratio_byRSA_plot <- plot_odds_ratio(avs_xclass_fisher_byRSA, "PDB_jury_column_rsa_unb")

patchworked_plot <- (feature_oddsratio_plot + feature_oddsratio_byRSA_plot
  + plot_layout(guides = 'collect', widths = c(1, 3))
  )


p_ranges_y <- 10^c(ggplot_build(patchworked_plot[[1]])$layout$panel_scales_y[[1]]$range$range,
                   ggplot_build(patchworked_plot[[2]])$layout$panel_scales_y[[1]]$range$range)

combined_plot <- (
    patchworked_plot
  & scale_y_log10(limits = c(1/1.6, 1.6),
                  breaks = c(0.7, 1, 1.5),
                  minor_breaks = c(0.6, 0.8, 0.9, 1.1, 1.2, 1.3, 1.4, 1.6)
  )
  & scale_x_discrete(limits = c('', 'cmd', 'cmn', 'cme', '', 'umd', 'umn', 'ume', ''))
  & theme(
    axis.line = element_line(color='black'),
    panel.grid.minor = element_line(size = 0.1),
    panel.grid.major = element_line(size = 0.1),
    panel.grid.major.x = element_blank()
    )
)

combined_plot
ggsave('plot4b.svg', combined_plot)
```
