---
title: "Figure 4C: Enrichment of Pathogenic variants in Conservation Plane"
output:
  github_document:
    toc: true
    toc_depth: 2
---

This notebook presents code that generates Figure 4C and supporting statistics,
which describe the enrichment of ClinVar Pathogenic variants within different
regions of the conservation plane.

The notebook works with our pre-calculated MSA column statistics for population
and clinical variants, and structural features, for relevant Pfams. If you want
to go straight to the plot code then follow the link in the Table of Contents above.

# Notebook resources
## Libraries

```{r Libraries}
suppressPackageStartupMessages(library(tidyverse))
library(patchwork)
```

## Dataset

```{r Load dataset}
# Read and process dataset
pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv <-
  read_csv("./data/pfam-gnomAD-clinvar-pdb-colstats_c7c3e19.csv.gz",
           comment = "#", show_col_types = FALSE)

avs <-
  pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv  # avs: alignments variants structure

# Filter unnecessary variables
whitelist <- c(
  "family",
  "column",
  "occupancy",
  "shenkin",
  "_missense_all",
  "mes_or",
  "mes_p",
  "occupancy_minus_threshold",
  "occupancy_gteq_threshold",
  "shenkin_nrank",
  "mes_or_nrank",
  "mes_or_log",
  "CV_Benign",
  "CV_Benign/Likely_benign",
  "CV_Conflicting_interpretations_of_pathogenicity",
  "CV_Likely_benign",
  "CV_Likely_pathogenic",
  "CV_Pathogenic",
  "CV_Uncertain_significance",
  "CV_Pathogenic/Likely_pathogenic",
  "PDB_jury_column_rsa_unb"
)
blacklist <- !names(avs) %in% whitelist
avs[, blacklist] <- NULL

# FYI: There are several other ClinVar entries I could look at in future
# "CV_not_provided",
# "CV_drug_response",
# "CV_association",
# "CV_risk_factor",
# "CV_other",
# "CV_Affects",
# "CV_protective",

# Better variable name format
names(avs)[names(avs) == '_missense_all'] <- 'missense_all'

avs <- (
  avs
  
  # Categorise MES and Shenkin
  |> mutate(mes_class = factor(
    ifelse(mes_p < 0.1, ifelse(mes_or < 1, 'md', 'me'), 'mn'), levels = c('md', 'mn', 'me')
  ))
  |> mutate(cons_class = factor(
    ifelse(shenkin_nrank > 0.5, 'u', 'c'), levels = c('u', 'c')
  ))
  |> mutate(xclass = factor(
    interaction(cons_class, mes_class, sep = ''),
    levels = c('cmd', 'cmn', 'cme', 'umd', 'umn', 'ume')
  ))
  
  # Other
  |> replace_na(list(CV_Pathogenic = 0))
)

# Clean-up
rm(pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv)


COLOUR_SCHEME <- c(
  'cmd' = '#CC79A7',
  'ume' = '#56B4E9',
  'umn' = '#999999',
  'cmn' = '#999999',
  'umd' = '#E69F00',
  'cme' = '#009E73'
)
```

# Calculate and plot odds ratios

1. Filter Pfams without any Pathogenic variants
2. Filter low-powered Pfams
3. Calculate summary statistics for each X-Class
4. Calculate odds ratios for each X-class
5. Repeat steps 3-4 for individual RSA classes
6. Compose plot

First we apply the filters:

```{r Filter dataset for these tests}
# Summarise counts
avs_test_set <- (
  avs
  
  # Column filters
  |> filter(occupancy > 0)
  
  # Pfam filters
  |> group_by(family) |> filter(any(CV_Pathogenic > 0)) |> ungroup()
  
  # Heuristic to select well-powered Pfams
  |> group_by(family) |> filter(any(mes_or < 1 & mes_p < 0.1)) |> ungroup()  # at least one depleted site
  |> group_by(family) |> filter(any(mes_or > 1 & mes_p < 0.1)) |> ungroup()  # at least one enriched site
)
```

Then summarise the data over the X-Classes:

```{r Summary statistics for different X-Classes}
summarize_data <- function(data, group_vars) {
  data |>
    group_by(across(all_of(group_vars))) |>
    summarise(
      n_pfams = n_distinct(family),
      n_columns = n(),
      n_residues = sum(occupancy),
      n_missense = sum(missense_all),
      across(c(CV_Pathogenic, CV_Benign), sum),
      cons_class = first(cons_class),
      mes_class = first(mes_class)
    )
}

add_ratios <-
  function(data, pathogenic_col, missense_col, residues_col) {
    data %>%
      mutate(
        missense_per_res = !!sym(missense_col) / !!sym(residues_col),
        patho_per_missense = !!sym(pathogenic_col) / !!sym(missense_col),
        patho_per_column = !!sym(pathogenic_col) / !!sym(residues_col)
      )
  }

add_test_backgrounds <- function(data, group_vars) {
  data |>
    ungroup() |>
    group_by(across(all_of(group_vars))) |>
    mutate(across(
      c(n_missense, n_residues, CV_Pathogenic, CV_Benign),
      sum,
      .names = "test_background_{.col}"
    ))
}


# Calculate group statistics for different X-Classes
avs_xclass_summary <- avs_test_set |>
  summarize_data(group_vars = "xclass") |>
  add_ratios(
    pathogenic_col = "CV_Pathogenic",
    missense_col = "n_missense",
    residues_col = "n_residues"
  ) |>
  add_test_backgrounds(group_vars = NULL)
avs_xclass_summary
```

```{r Totals}
# Totals
avs_xclass_summary |> ungroup() |> select(where(is.numeric)) |> summarise_all(sum)
```

## Odds ratios for Pathogenic enrichment in X-Classes
Now apply Fishers exact test for each X-Class (CMD, UMD, etc...).

```{r Test enrichment of protein-ligand interactions in different X-Classes}
perform_fisher_test <- function(df, group_vars) {
  df %>%
    group_by(!!!syms(group_vars)) %>%
    summarise(
      test = broom::tidy(
        fisher.test(matrix(
          c(
            CV_Pathogenic,
            n_missense,
            test_background_CV_Pathogenic - CV_Pathogenic,
            test_background_n_missense - n_missense
          ),
          nrow = 2
        ), alternative = 'two.sided')
      ),
      cons_class = first(cons_class),
      mes_class = first(mes_class)
    ) %>%
    unnest(test)
}

avs_xclass_fisher <- perform_fisher_test(avs_xclass_summary, "xclass")
avs_xclass_fisher
```

Now do the same while stratifying for RSA class as well.

```{r Summarise data}
avs_xclass_summary_byRSA <- avs_test_set |>
  filter(PDB_jury_column_rsa_unb %in% c('Core', 'Part. Exposed', 'Surface')) |>
  summarize_data(group_vars = c("PDB_jury_column_rsa_unb", "xclass")) |>
  add_ratios(
    pathogenic_col = "CV_Pathogenic",
    missense_col = "n_missense",
    residues_col = "n_columns"
  ) |>
  add_test_backgrounds(group_vars = "PDB_jury_column_rsa_unb")
avs_xclass_summary_byRSA
```

```{r Totals by RSA}
# Totals
avs_xclass_summary_byRSA |> ungroup() |> select(where(is.numeric)) |> summarise_all(sum)
```

## Odds ratios for Pathogenic enrichment in X-Class * RSA

```{r Test enrichment of Pathogenic variants in X-Classes by RSA}
avs_xclass_fisher_byRSA <- perform_fisher_test(avs_xclass_summary_byRSA, c("PDB_jury_column_rsa_unb", "xclass"))
avs_xclass_fisher_byRSA
```

# Figure 4C: Odds ratio plot

```{r Combined plot}
plot_odds_ratio <- function(df, facet_var = NULL) {
  dodge_width <- 0.6
  
  plot <- df |>
    ggplot() +
    aes(x = xclass, y = estimate, colour = xclass) +
    geom_hline(yintercept = 1, lty = 2) +
    geom_errorbar(
      aes(ymin = sapply(conf.low, function(y) max(y, 0.2)),  # clip
          ymax = conf.high),
      width = 0.5,
      position = position_dodge(dodge_width)
    ) +
    geom_point(position = position_dodge(dodge_width)) +
    scale_y_log10() +
    ylab('Odds Ratio') +
    xlab('') +
    theme_minimal() +
    theme(legend.position = "none", aspect.ratio = 1) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) +
    scale_color_manual(values = COLOUR_SCHEME)
  
  if (!is.null(facet_var)) {
    plot <- plot + facet_grid(paste0(" ~ ", facet_var))
  }
  
  return(plot)
}


feature_oddsratio_plot <- plot_odds_ratio(avs_xclass_fisher)
feature_oddsratio_byRSA_plot <- plot_odds_ratio(avs_xclass_fisher_byRSA, "PDB_jury_column_rsa_unb")

patchworked_plot <- (feature_oddsratio_plot + feature_oddsratio_byRSA_plot
  + plot_layout(guides = 'collect', widths = c(1, 3))
  )


p_ranges_y <- 10^c(ggplot_build(patchworked_plot[[1]])$layout$panel_scales_y[[1]]$range$range,
                   ggplot_build(patchworked_plot[[2]])$layout$panel_scales_y[[1]]$range$range)

combined_plot <- (
    patchworked_plot
  & scale_y_log10(limits = c(0.2, 6),  # limit in conjunction with clip
                  breaks = c(0.1, 0.2, 0.3, 0.4, 0.5, 1, 2, 3, 4, 5, 6),
                  labels = c('', '', 0.3, '', 0.5, 1, 2, 3, 4, 5, ''),
                  minor_breaks = c(seq(0.1, 1, 0.1), seq(1, 5, 1)),
  )
  & scale_x_discrete(limits = c('', 'cmd', 'cmn', 'cme', '', 'umd', 'umn', 'ume', ''))
  & theme(
    axis.line = element_line(color='black'),
    panel.grid.minor = element_line(size = 0.1),
    panel.grid.major = element_line(size = 0.1),
    panel.grid.major.x = element_blank()
    )
)

combined_plot
ggsave('plot4c.svg', combined_plot)
```

# Supporting statistics

We quoted the maximum discrimination observed with respect to X-Class and
ClinVar Pathogenic variants. This is found by comparing Surface CMD to UME
positions:

```{r Surface CMD vs. UME}
(
  avs
  |> filter(PDB_jury_column_rsa_unb == 'Surface')
  |> filter(xclass %in% c('cmd', 'ume'))
  |> group_by(xclass)
  |> summarise(
    n_patho = sum(CV_Pathogenic),
    n_missense = sum(missense_all)
  )
  -> cmd_vs_ume
)

(
  cmd_vs_ume
  |> pivot_wider(names_from = xclass, values_from = starts_with("n_"))
  |> summarise(
      test =  broom::tidy(
        fisher.test(matrix(c(n_patho_cmd, n_missense_cmd, n_patho_ume, n_missense_ume),
                           nrow = 2), alternative = 'two.sided'))
    )
  |> unnest(test)
)
```

It's sensible to check that we observe the widely reported enrichment of
Pathogenic variants in buried positions:

```{r Positive control: do we see Pathogenic variant senriched in Core?}
avs |>
  group_by(PDB_jury_column_rsa_unb) |>
  summarise(n_missense = sum(missense_all), CV_Pathogenic = sum(CV_Pathogenic)) |>
  mutate(across(c(n_missense, CV_Pathogenic), sum, .names = "test_background_{.col}")) |>
  group_by(PDB_jury_column_rsa_unb, .drop = TRUE) |>
  summarise(
    # Biolip enrichment
    test =  broom::tidy(
      fisher.test(matrix(c(CV_Pathogenic, n_missense,
                           test_background_CV_Pathogenic - CV_Pathogenic,
                           test_background_n_missense - n_missense),
                         nrow = 2), alternative = 'two.sided'))
  ) |>
  unnest(test)
```

# Future work

-   ClinVar enrichments in different positions with various structural features
