---
title: "Table 1: The Conservation Plane"
output:
  github_document:
    toc: true
    toc_depth: 2
---

# Introduction
This R markdown document tabulates summary data for the different conservation
plane categories. The dataset used for this analysis is obtained from our
pre-calculated MSA column statistics for Pfam, included in this repository. The
output includes data used in Table 1 in the manuscript.

# Notebook resources
## Libraries

```{r Libraries}
suppressPackageStartupMessages(library(tidyverse))
```

## Dataset

```{r Load dataset}
# Read and process dataset
pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv <-
  read_csv(
    "data/pfam-gnomAD-clinvar-pdb-colstats_c7c3e19.csv.gz",
    comment = "#",
    show_col_types = FALSE
  )

# Processing
avs <-
  pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv  # avs: alignments variants structure

# Filter unnecessary variables
whitelist <- c(
  "family",
  "column",
  "splice_region_variant&synonymous_variant",
  "synonymous_variant",
  "occupancy",
  "shenkin",
  "_missense_all",
  "mes_or",
  "mes_p",
  "occupancy_minus_threshold",
  "occupancy_gteq_threshold",
  "shenkin_nrank",
  "mes_or_nrank",
  "mes_or_log",
  "CV_Pathogenic",
  "PDB_sequences_with_contacts",
  "PDB_jury_column_rsa_unb"
)
blacklist <- !names(avs) %in% whitelist
avs[, blacklist] <- NULL

# Better variable name format
names(avs)[names(avs) == '_missense_all'] <- 'missense_all'

# Calculate synonymous all
avs[, 'synonymous_all'] <-
  avs[, 'splice_region_variant&synonymous_variant'] +
  avs[, 'synonymous_variant']

# Clean-up
rm(pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv)
```

Below is where I define the conservation plane categories.

```{r Categorise columns}
# Apply thresholds to categorise MES and Shenkin and the combined class
avs <- (
  avs
  |> mutate(mes_class = factor(
    ifelse(mes_p < 0.1, ifelse(mes_or < 1, 'md', 'me'), 'mn'), levels = c('md', 'mn', 'me')
  ))
  |> mutate(cons_class = factor(
    ifelse(shenkin_nrank > 0.5, 'u', 'c'), levels = c('u', 'c')
  ))
  |> mutate(xclass = factor(
    interaction(cons_class, mes_class, sep = ''),
    levels = c('cmd', 'cmn', 'cme', 'umd', 'umn', 'ume')
  ))
)
```

```{r Standard filters}
# Filters
apply_filters <- function(df,
                          occupancy = FALSE,
                          has_RSA = FALSE,
                          has_depleted = FALSE,
                          has_enriched = FALSE,
                          has_PDB = FALSE,
                          has_Pathogenic = FALSE) {
  if (occupancy)
    df <- df |> filter(occupancy > 0)  # must have human sequences
  if (has_RSA)
    df <-
      df |> filter(PDB_jury_column_rsa_unb %in% c('Core', 'Part. Exposed', 'Surface'))
  
  # Heuristic for well-powered Pfams
  if (has_depleted)
    df <-
      df |>
      group_by(family) |>
      filter(any(mes_or < 1 & mes_p < 0.1)) |> ungroup()  # >1 depleted
  if (has_enriched)
    df <-
      df |>
      group_by(family) |>
      filter(any(mes_or > 1 & mes_p < 0.1)) |> ungroup()  # >1 enriched
  
  # Has PDB data
  if (has_PDB)
    df <-
      df |>
      group_by(family) |>
      filter(any(PDB_sequences_with_contacts >= 1)) |> ungroup()
  
  # Has ClinVar Pathogenic
  if (has_Pathogenic)
    df <- df |> group_by(family) |> filter(any(CV_Pathogenic > 0)) |> ungroup()
  return(df)
}
```

# Table 1: The Conservation Plane (table data)
This table reports summary statistics for each category defined by the
classification of Pfam domains by population missense constraint and
evolutionary divergence.

```{r Summarise by xclass}
avs |>
  apply_filters(occupancy = TRUE,
                has_RSA = FALSE,
                has_depleted = TRUE,
                has_enriched = FALSE,
                has_PDB = FALSE,
                has_Pathogenic = FALSE) |>
  group_by(xclass) |>
  summarise(Families = n_distinct(family),
            Columns = n(),
            Residues = sum(occupancy),
            Missense = sum(missense_all)) ->
  avs_summary
avs_summary
```
```{r Totals}
avs_summary |>
  summarise(Families = max(Families),
            across(where(is.numeric), ~ sum(.x, na.rm = TRUE))) ->
  avs_summary_totals
avs_summary_totals
```


```{r Save datafiles, include=FALSE}
write_csv(avs_summary, 'Table1_Summary-statistics-for-X-classes.csv')
write_csv(avs_summary_totals, 'Table1-Totals-Row_Summary-statistics-for-X-classes.csv')
```

