---
title: "Figure 2A: RSA class stratified MES and Shenkin residue ECDFs"
output:
  github_document:
    toc: true
    toc_depth: 2
---

# Introduction
This R markdown document visualizes relationships between solvent accessibility
with both evolutionary conservation and population constraint in Pfam domains.

The dataset used for this analysis is obtained from our pre-calculated MSA
column statistics for Pfam, included in this repository. The resulting plots
corresponds to Figure 2A in the manuscript.

# Notebook resources
## Libraries

```{r Dependencies}
suppressPackageStartupMessages(library(tidyverse))
```

## Dataset

```{r Load dataset}

# Load pfam-gnomAD dataset}
pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv <-
  read_csv("data/pfam-gnomAD-clinvar-pdb-colstats_c7c3e19.csv.gz", comment = "#")

# Processing
avs <- pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv  # avs: alignments variants structure

# Filter columns
whitelist <- c(
  "family",
  "column",
  "occupancy",
  "shenkin",
  "_missense_all",
  "mes_or",
  "mes_p",
  "occupancy_minus_threshold",
  "occupancy_gteq_threshold",
  "shenkin_nrank",
  "mes_or_nrank",
  "mes_or_log",
  "PDB_sequences_with_contacts",
  "PDB_jury_column_rsa_unb",
  "PDB_jury_column_rsa_bound"
)

blacklist <- !names(avs) %in% whitelist
avs[, blacklist] <- NULL

# Simplify column names
names(avs)[names(avs) == '_missense_all'] <- 'missense_all'
```

```{r Categorise columns}
# Apply thresholds to categorise MES and Shenkin and the combined class
avs <- (
  avs
  |> mutate(mes_class = factor(
    ifelse(mes_p < 0.1, ifelse(mes_or < 1, 'md', 'me'), 'mn'), levels = c('md', 'mn', 'me')
  ))
  |> mutate(cons_class = factor(
    ifelse(shenkin_nrank > 0.5, 'u', 'c'), levels = c('u', 'c')
  ))
  |> mutate(xclass = factor(
    interaction(cons_class, mes_class, sep = ''),
    levels = c('cmd', 'cmn', 'cme', 'umd', 'umn', 'ume')
  ))
)
```
 
``` {r Minimal filters and minor data cleaning}
(avs <- avs
  # Column filters
  |> filter(occupancy > 0)  # must have human sequences
  # |> filter(PDB_sequences_with_contacts > 0)  # drop Pfams with no PDB data?

  # Powered Pfams
  |> group_by(family) |> filter(any(mes_or < 1 & mes_p < 0.1)) |> ungroup()
  |> group_by(family) |> filter(any(mes_or > 1 & mes_p < 0.1)) |> ungroup()
  
  # Other
  |> mutate(PDB_jury_column_rsa_unb = as.factor(PDB_jury_column_rsa_unb))  # count all RSA*MES intersections
  |> replace_na(list(PDB_protein_ligand_interactions = 0, PDB_biolip = 0,
                     PDB_sequences_with_contacts = 0))
)

rm(pfam_gnomAD_clinvar_pdb_colstats_c7c3e19_csv)
```


```{r Exposure class group sizes}
 RSA_LEVELS <- c('Core', 'Part. Exposed', 'Surface')

# N residues per exposure with high/low conservation
(avs
 |> filter(PDB_jury_column_rsa_unb %in% c('Core', 'Part. Exposed', 'Surface'))
 |> group_by(shenkin_nrank < 0.5, PDB_jury_column_rsa_unb)
 |> summarise(n())
 )

# N residues per exposure with high/low MES
(avs
 |> filter(PDB_jury_column_rsa_unb %in% c('Core', 'Part. Exposed', 'Surface'))
 |> group_by(mes_or < 1, PDB_jury_column_rsa_unb)
 |> summarise(n())
 )
```


# Figure 2A: Residue Shenking/MES ECDFs for each exposure class (manuscript figure)

```{r Figure 2A (right panel): ECDF of residue Shenkin for different exposure classes}
COLOUR_SCHEME <- c(
  'partex_cum_frac' = '#56B4E9',
  'core_cum_frac' = '#E69F00',
  'surface_cum_frac' = '#009E73'
)

(avs
  |> filter(PDB_jury_column_rsa_unb %in% RSA_LEVELS)
  |> filter(occupancy_gteq_threshold)
  |> filter(occupancy > 0)
  |> filter(occupancy >= 20)
  # |> filter(occupancy < 50). # variation with bounded occupancy window
  |> arrange(shenkin_nrank)
  |> select(family, column, shenkin_nrank, PDB_jury_column_rsa_unb)
  |> pivot_wider(names_from = PDB_jury_column_rsa_unb,
                 values_from = PDB_jury_column_rsa_unb,
                 values_fn = function(x) return(1),
                 values_fill = 0)
  |> mutate(core_cum_frac = cumsum(Core) / sum(Core),
            partex_cum_frac = cumsum(`Part. Exposed`) / sum(`Part. Exposed`),
            surface_cum_frac = cumsum(Surface) / sum(Surface))
  |> pivot_longer(core_cum_frac:surface_cum_frac)
  |> ggplot()
  + aes(x = shenkin_nrank, y = value*100, colour = name)
  + geom_abline(slope = 100, intercept = 0, lty = 2)
  + geom_line()
  
  # annotation
  + xlab('Normalised Shenkin')
  + ylab('Proportion of positions (%)')

  # plot theme
  + scale_colour_manual(values = COLOUR_SCHEME)
  + theme_minimal()
  + theme(aspect.ratio = 1)
  + theme(
    axis.line = element_line(color='black'),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(size = 0.1),
    legend.position = "none"
  )
)

ggsave('plot2a_right.svg')
# ggsave('plot2a_right_supp-occ200.svg')
```


```{r Figure 2A (left panel): ECDF of residue MES for different exposure classes}

(avs
  |> filter(PDB_jury_column_rsa_unb %in% RSA_LEVELS)
  |> filter(occupancy_gteq_threshold)
  |> filter(occupancy > 0)
  |> filter(occupancy >= 20)
  |> arrange(mes_or_nrank)
  |> select(family, column, mes_or_nrank, PDB_jury_column_rsa_unb)
  |> pivot_wider(names_from = PDB_jury_column_rsa_unb,
                 values_from = PDB_jury_column_rsa_unb,
                 values_fn = function(x) return(1),
                 values_fill = 0)
  |> mutate(core_cum_frac = cumsum(Core) / sum(Core),
            partex_cum_frac = cumsum(`Part. Exposed`) / sum(`Part. Exposed`),
            surface_cum_frac = cumsum(Surface) / sum(Surface))
  |> pivot_longer(core_cum_frac:surface_cum_frac)
  |> ggplot()
  + aes(x = mes_or_nrank, y = value*100, colour = name)
  + geom_abline(slope = 100, intercept = 0, lty = 2)
  + geom_line()
 
   # annotation
  + xlab('Normalised MES')
  + ylab('Proportion of positions (%)')

  # plot theme
  + scale_colour_manual(values = COLOUR_SCHEME)
  + theme_minimal()
  + theme(aspect.ratio = 1)
  + theme(
    axis.line = element_line(color='black'),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(size = 0.1),
    legend.position = "none"
  )
)

ggsave('plot2a_left.svg')
# ggsave('plot2a_left_supp-occ200.svg')
```

# Calculating some quoted figures in the manuscript

```{r Quoted figures - MES}
(
avs
  |> filter(PDB_jury_column_rsa_unb %in% RSA_LEVELS)
  |> filter(occupancy_gteq_threshold)
  |> filter(occupancy > 0)
  |> filter(occupancy > 20)
  |> arrange(mes_or_nrank)
  |> select(family, column, mes_or_nrank, PDB_jury_column_rsa_unb)
  |> pivot_wider(names_from = PDB_jury_column_rsa_unb,
                 values_from = PDB_jury_column_rsa_unb,
                 values_fn = function(x) return(1),
                 values_fill = 0)
  |> mutate(core_cum_frac = cumsum(Core) / sum(Core),
            partex_cum_frac = cumsum(`Part. Exposed`) / sum(`Part. Exposed`),
            surface_cum_frac = cumsum(Surface) / sum(Surface))
  |> pivot_longer(core_cum_frac:surface_cum_frac)
  |> filter(mes_or_nrank >= 0.5)
  |> group_by(name)
  |> summarise_all(min)
)
```

```{r Quoted figures - Shenkin}
(avs
  |> filter(PDB_jury_column_rsa_unb %in% RSA_LEVELS)
  |> filter(occupancy_gteq_threshold)
  |> filter(occupancy > 0)
  |> filter(occupancy >= 20)
  # |> filter(occupancy < 50)
  |> arrange(shenkin_nrank)
  |> select(family, column, shenkin_nrank, PDB_jury_column_rsa_unb)
  |> pivot_wider(names_from = PDB_jury_column_rsa_unb,
                 values_from = PDB_jury_column_rsa_unb,
                 values_fn = function(x) return(1),
                 values_fill = 0)
  |> mutate(core_cum_frac = cumsum(Core) / sum(Core),
            partex_cum_frac = cumsum(`Part. Exposed`) / sum(`Part. Exposed`),
            surface_cum_frac = cumsum(Surface) / sum(Surface))
  |> pivot_longer(core_cum_frac:surface_cum_frac)
  |> filter(shenkin_nrank >= 0.5)
  |> group_by(name)
  |> summarise_all(min)
)
```


```{r Session Info.}
sessionInfo()
```

